Recursive Solutions to First-Order Model Counting

### 2022-02-17 notes/notes.tex
Description of smoothing

\usetikzlibrary{arrows,fit,positioning,trees,cd}

@@

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable, clause, unit clause
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item maybe: notation for partial function, notation for powerset, notation for disjoint union, domain size
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\item notation for variable substitution for literal and constraint sets: $S[x/V]$, where $S$ is the set, $V$ is a set of variable, and $x$ is the new constant or variable. Note that for constraint sets this could mean that a variable and a constant need to switch places.
\item During inference, there is a domain size map $\size\colon \mathscr{D} \to \mathbb{N}_0$.
\item $\sqcup$ for both sets and functions
\end{itemize}

\paragraph{TODO}
\begin{itemize}
\item should I always say `compilation rule' instead of `operator'?
\item maybe mathcal instead of mathscr
\item maybe $\pi$ is global enough to have a more unique name
\item maybe $c[\beta, \gamma]$ is better than $c \ast (\beta, \gamma)$
\item capitalise variable names.
\item introduce/use notation for vertices, edges of a graph
\end{itemize}

@@

\section{Search}

\begin{definition}
  A \emph{first-order deterministic decomposable negation normal form computational graph} (FCG) is a (weakly connected) directed graph with a a single source, node labels, and ordered direct successors. We denote an FCG as $G = (V, s, N^+)$, where $V$ is the set of nodes, $s \in V$ is the source, and $N^+$ is the direct successor function that maps each node in $V$ to a list that contains either other nodes in $V$ or $\star$, which means that the target of the edge is yet to be determined. For each $v \in V$, the length of the list $N^+(v)$ is determined by the label of $v$.

@@

\paragraph{TODO.}
\begin{itemize}
\item maybe add a (vertex) labelling function and its codomain (and use it in the algorithm)? But it's only used once...
\item explain the connection to the algebraic formalism (or just don't use it). % We will also use algebraic notation to describe (parts of) circuits: circuittype_parameters(nodes pointed to by the outedges)
\item define `an FCG for a formula'
\item vertex or node, $v$ or $n$?
\item what notation for lists (a.k.a. sequences) am I going to use? $\oplus$ for concatenation (of both two lists and an element to a list), $\in$ for (in-order) enumeration, $[]$ for an empty list, $[x]$ for a list with one element, $|L|$ for the length of the list, and $h : t$ to denote a list with first element (a.k.a. head) $h$ and remaining list (a.k.a. tail) $t$.
\item use Greek letters for compilation rules
\end{itemize}

@@

\begin{algorithm}
  \caption{Functions used in \cref{alg:search} for applying compilation rules}
  \label{alg:apply}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwData{newStates}{newStates}
  \SetKwFunction{applyGreedyRules}{applyGreedyRules}
  \SetKwFunction{applyNonGreedyRules}{applyNonGreedyRules}
  \SetKwFunction{applyGreedyRulesToAllFormulas}{applyGreedyRulesToAllFormulas}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \KwData{a set of greedy rules $\Gamma$}
  \KwData{a set of non-greedy rules $\Delta$}
  \Fn{\applyGreedyRules{$\phi$, $C$}}{
    \ForEach{rule $r \in \Gamma$}{
      $S \gets r(\phi)$\;
      \If{$S \ne \emptyset$}{
        $(G, L) \gets \text{any element of } S$\;
        \lIf{$G = {\normalfont \texttt{null}}$}{\Return{\applyGreedyRules{the only element of $L$, $C$}}}
        \Else{
          $(V, s, N^+) \gets G$\;
          $C \gets \updateCache(C, \phi, s)$\;
          \Return{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}\;
        }
      }
    }
    \Return{$({\normalfont \texttt{null}}, C, [\phi])$}\;
  }
  \Fn{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}{
    $(V, s, N^+) \gets G$\;
    $N^+(s) \gets []$\;
    $L' \gets []$\;
    \ForEach{formula $\phi \in L$}{
      $(G', C, L'') \gets \applyGreedyRules{$\phi$, $C$}$\;
      $L' \gets L' \oplus L''$\;
      \lIf{$G' = {\normalfont \texttt{null}}$}{$N^+(s) \gets N^+(s) \oplus [\star]$}
      \Else{
        $(V', s', N') \gets G'$\;
        $V \gets V \sqcup V'$\;
        $N^+ \gets N^+ \sqcup N'$\;
        $N^+(s) \gets N^+(s) \oplus [s']$\;
      }
    }
    \Return{$((V, s, N^+), C, L')$}\;
  }
  \Fn{\applyNonGreedyRules{$s$}}{
    $(G, C, L) \gets s$\;
    $\phi : T \gets L$\;
    $(G', C', L') \gets \text{a copy of } s$\;
    $\newStates \gets []$\;
    \ForEach{rule $r \in \Delta$}{
      \ForEach{$(G'', L'') \in r(\phi)$}{
        \lIf{$G'' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus \applyNonGreedyRules{$(G', C', L'')$}$}
        \Else{
          $(V, s, N^+) \gets G''$\;
          $C' \gets \updateCache{$C'$, $\phi$, $s$}$\;
          $(G'', C', L'') \gets \applyGreedyRulesToAllFormulas{$G''$, $C'$, $L''$}$\;
          \lIf{$G' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus [(G'', C', L'' \oplus T)]$}
          \lElse{$\newStates \gets \newStates \oplus [(\mergeFcgs{$G'$, $G''$}, C', L'' \oplus T)]$}
        }
      }
      $(G', C', L') \gets \text{a copy of } s$\;
    }
    \Return{\newStates}\;
  }
\end{algorithm}

\begin{algorithm}
  \caption{Helper functions used by \cref{alg:apply}}
  \label{alg:helpers}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \Fn{\updateCache{$C$, $\phi$, $v$}}{
    \lIf{$\#\phi \not\in \dom(C)$}{\Return{$C \cup \{\, \#\phi \mapsto (\phi, v) \,\}$}}
    \lIf{there is no $(\phi', v') \in C(\#\phi)$ such that $v' = v$}{$C(\#\phi) \gets (\phi, v) \oplus C(\#\phi)$}
    \Return{$C$}\;
  }
  \Fn{\mergeFcgs{$G = (V, s, N^+)$, $G' = (V', s', N')$, $r = s$}}{
    \lIf{$r$ has label \textsc{Ref}}{\Return{\normalfont \texttt{null}}}
    \ForEach{$t \in N^+(r)$}{
      \If{$t = \star$}{
        replace $t$ with $s'$ in $N^+(r)$\;
        \Return{$(V \sqcup V', s, N^+ \sqcup N')$}\;
      }
      $G'' \gets \mergeFcgs{$G$, $G'$, $t$}$\;
      \lIf{$G'' \ne {\normalfont \texttt{null}}$}{\Return{$G''$}}
    }
    \Return{\normalfont \texttt{null}}\;
  }
\end{algorithm}

@@

@@

\begin{itemize}
\item domains, smoothing, and avoiding infinite cycles
\item new rules that don't create nodes (e.g., duplicate removal, unconditional contradiction detection, etc.)
\item my search algorithm
\item some notes on halting
  \begin{itemize}
  \item Search is infinite. Some rules increase the size of the formula(s), but most reduce it.
  \item Inference is guaranteed to terminate if at least one domain shrinks by at least one. But note that allowing recursive calls with the same domain sizes (e.g., $f(n) = f(n) + \dots$) could be useful because these problematic terms might cancel out.
  \item It's impossible for $n \gets n - 1$ and $\texttt{for } n \in \dots$ to combine in a way that results in an infinite loop.
  \end{itemize}
\end{itemize}

@@

TODO: define 'circuit' (or something similar). Note that circuits are rooted.

>>>>

\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[triangle/.style = {regular polygon, regular polygon sides=3}]
    \node[draw,ellipse] (a) at (0, 0) {$\phi(n)$};
    \node[draw,ellipse] (b) at (0, -2) {$\psi(n-1)$};
    \node[draw,ellipse] (c) at (-1, -4) {$\phi(n-1)$};
    \node[draw,triangle] (d) at (1, -4) {};
    \draw[-{Latex[length=2mm]}] (a) -- (b) node [midway,xshift=50] {domain recursion};
    \draw[-{Latex[length=2mm]}] (b) -- (c);
    \draw[-{Latex[length=2mm]}] (b) -- (d);
    \draw[-{Latex[length=2mm]}] (c) to [bend left=45] node [midway,xshift=-30] () {$n \mapsto n-1$} (a);
  \end{tikzpicture}
  \caption{An illustration of the main idea. TODO: refer to this in the introduction.}
\end{figure}

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable, clause, unit clause
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item maybe: notation for partial function, notation for powerset, notation for disjoint union, domain size
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\item notation for variable substitution for literal and constraint sets: $S[x/V]$, where $S$ is the set, $V$ is a set of variable, and $x$ is the new constant or variable. Note that for constraint sets this could mean that a variable and a constant need to switch places.
\item During inference, there is a domain size map $\size\colon \mathscr{D} \to \mathbb{N}_0$.
\item $\sqcup$ for both sets and functions
\item mention which rules are in $\Gamma$ and which ones are in $\Delta$ (and why tryCache has to be in $\Delta$).
\end{itemize}

\paragraph{TODO}
\begin{itemize}
\item should I always say `compilation rule' instead of `operator'?
\item maybe mathcal instead of mathscr
\item maybe $\pi$ is global enough to have a more unique name
\item maybe $c[\beta, \gamma]$ is better than $c \ast (\beta, \gamma)$
\item capitalise variable names.
\item introduce/use notation for vertices, edges of a graph
\item maybe `plus-plus' is better than $\oplus$
\item maybe put each function into its own algorithm environment, using the caption for the function name
\end{itemize}

@@

\section{Search/Compilation}

\begin{definition}
  A \emph{first-order deterministic decomposable negation normal form computational graph} (FCG) is a (weakly connected) directed graph with a a single source, node labels, and ordered direct successors. We denote an FCG as $G = (V, s, N^+, \lambda)$, where $V$ is the set of nodes, $s \in V$ is the source, $N^+$ is the direct successor function that maps each node in $V$ to a list that contains either other nodes in $V$ or $\star$, which means that the target of the edge is yet to be determined, and $\lambda\colon V \to \mathscr{L}$ is the vertex-labelling function with some codomain $\mathscr{L}$, which is assumed to be the same for all FCGs. For each $v \in V$, the length of the list $N^+(v)$ is determined by $\lambda(v)$.

@@

\paragraph{TODO.}
\begin{itemize}
\item explain the connection to the algebraic formalism (or just don't use it). % We will also use algebraic notation to describe (parts of) circuits: circuittype_parameters(nodes pointed to by the outedges)
\item define `an FCG for a formula'
\item vertex or node, $v$ or $n$?
\item what notation for lists (a.k.a. sequences) am I going to use? $\oplus$ for concatenation (of both two lists and an element to a list), $\in$ for (in-order) enumeration, $[]$ for an empty list, $[x]$ for a list with one element, $|L|$ for the length of the list, and $h : t$ to denote a list with first element (a.k.a. head) $h$ and remaining list (a.k.a. tail) $t$. Also add an example of the list comprehension notation.
\item use Greek letters for compilation rules
\end{itemize}

@@

\begin{algorithm}
  \caption{Functions used in \cref{alg:search} for applying compilation rules}
  \label{alg:apply}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwData{newStates}{newStates}
  \SetKwFunction{applyGreedyRules}{applyGreedyRules}
  \SetKwFunction{applyNonGreedyRules}{applyNonGreedyRules}
  \SetKwFunction{applyGreedyRulesToAllFormulas}{applyGreedyRulesToAllFormulas}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \KwData{a set of greedy rules $\Gamma$}
  \KwData{a set of non-greedy rules $\Delta$}
  \Fn{\applyGreedyRules{$\phi$, $C$}}{
    \ForEach{rule $r \in \Gamma$}{
      $S \gets r(\phi)$\;
      \If{$S \ne \emptyset$}{
        $(G, L) \gets \text{any element of } S$\;
        \lIf{$G = {\normalfont \texttt{null}}$}{\Return{\applyGreedyRules{the only element of $L$, $C$}}}
        \Else{
          $(V, s, N^+, \lambda) \gets G$\;
          $C \gets \updateCache(C, \phi, s)$\;
          \Return{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}\;
        }
      }
    }
    \Return{$({\normalfont \texttt{null}}, C, [\phi])$}\;
  }
  \Fn{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}{
    $(V, s, N^+, \lambda) \gets G$\;
    $N^+(s) \gets []$\;
    $L' \gets []$\;
    \ForEach{formula $\phi \in L$}{
      $(G', C, L'') \gets \applyGreedyRules{$\phi$, $C$}$\;
      $L' \gets L' \oplus L''$\;
      \lIf{$G' = {\normalfont \texttt{null}}$}{$N^+(s) \gets N^+(s) \oplus [\star]$}
      \Else{
        $(V', s', N', \lambda') \gets G'$\;
        $V \gets V \sqcup V'$\;
        $N^+ \gets N^+ \sqcup N'$\;
        $N^+(s) \gets N^+(s) \oplus [s']$\;
        $\lambda \gets \lambda \sqcup \lambda'$\;
      }
    }
    \Return{$((V, s, N^+, \lambda), C, L')$}\;
  }
  \Fn{\applyNonGreedyRules{$s$}}{
    $(G, C, L) \gets s$\;
    $\phi : T \gets L$\;
    $(G', C', L') \gets \text{a copy of } s$\;
    $\newStates \gets []$\;
    \ForEach{rule $r \in \Delta$}{
      \ForEach{$(G'', L'') \in r(\phi)$}{
        \lIf{$G'' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus \applyNonGreedyRules{$(G', C', L'')$}$}
        \Else{
          $(V, s, N^+, \lambda) \gets G''$\;
          $C' \gets \updateCache{$C'$, $\phi$, $s$}$\;
          $(G'', C', L'') \gets \applyGreedyRulesToAllFormulas{$G''$, $C'$, $L''$}$\;
          \lIf{$G' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus [(G'', C', L'' \oplus T)]$}
          \lElse{$\newStates \gets \newStates \oplus [(\mergeFcgs{$G'$, $G''$}, C', L'' \oplus T)]$}
        }
      }
      $(G', C', L') \gets \text{a copy of } s$\;
    }
    \Return{\newStates}\;
  }
\end{algorithm}

\begin{algorithm}
  \caption{Helper functions used by \cref{alg:apply}}
  \label{alg:helpers}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \Fn{\updateCache{$C$, $\phi$, $v$}}{
    \lIf{$\#\phi \not\in \dom(C)$}{\Return{$C \cup \{\, \#\phi \mapsto (\phi, v) \,\}$}}
    \lIf{there is no $(\phi', v') \in C(\#\phi)$ such that $v' = v$}{$C(\#\phi) \gets (\phi, v) \oplus C(\#\phi)$}
    \Return{$C$}\;
  }
  \Fn{\mergeFcgs{$G = (V, s, N^+, \lambda)$, $G' = (V', s', N', \lambda')$, $r = s$}}{
    \lIf{$\lambda(r) = \textsc{Ref}$}{\Return{\normalfont \texttt{null}}}
    \ForEach{$t \in N^+(r)$}{
      \If{$t = \star$}{
        replace $t$ with $s'$ in $N^+(r)$\;
        \Return{$(V \sqcup V', s, N^+ \sqcup N', \lambda \sqcup \lambda')$}\;
      }
      $G'' \gets \mergeFcgs{$G$, $G'$, $t$}$\;
      \lIf{$G'' \ne {\normalfont \texttt{null}}$}{\Return{$G''$}}
    }
    \Return{\normalfont \texttt{null}}\;
  }
\end{algorithm}

@@

\section{Smoothing}

\begin{algorithm}
  \SetKwData{changed}{changed}
  \caption{Propagating atoms for smoothing across the FCG in a way that avoids infinite loops}
  \label{alg:smoothing}
  \KwIn{FCG $(V, s, N^+, \lambda)$}
  \KwIn{function $\iota$ that maps labels in $\mathscr{L}$ to sets of atoms}
  \KwIn{functions $\{\,f_l\,\}_{l \in \mathscr{L}}$ that take a list of sets of atoms and return a set of atoms}
  \KwOut{function $S$ that maps vertices in $V$ to sets of atoms}
  $S \gets \{\, v \mapsto \iota(\lambda(v)) \mid v \in V \,\}$\;
  $\changed \gets \texttt{true}$\;
  \While{\changed}{
    $\changed \gets \texttt{false}$\;
    \ForEach{vertex $v \in V$}{
      $S' \gets f_{\lambda(v)}([S(w) \mid w \in N^+(v)])$\;
      \If{$S' \ne S(v)$}{
        $\changed \gets \texttt{true}$\;
        $S(v) \gets S'$\;
      }
    }
  }
\end{algorithm}

[Insert motivation for smoothing from Section 3.4. of the Forclift paper.] Originally, smoothing was (and still is) a two-step process. First, atoms that are still accounted for in the circuit are propagated upwards. Then, at nodes with certain labels, missing atoms are detected and additional sinks are created to account for them. If left unchanged, the first step of this process would result in an infinite loop whenever a cycle is encountered. \Cref{alg:smoothing} outlines how the first step can be adapted to an arbitrary directed graph.

@@

\begin{itemize}
\item new rules that don't create nodes (e.g., duplicate removal, unconditional contradiction detection, etc.)
\item some notes on halting
  \begin{itemize}
  \item Search is infinite. Some rules increase the size of the formula(s), but most reduce it.
  \item Inference is guaranteed to terminate if at least one domain shrinks by at least one. But note that allowing recursive calls with the same domain sizes (e.g., $f(n) = f(n) + \dots$) could be useful because these problematic terms might cancel out.
  \item It's impossible for $n \gets n - 1$ and $\texttt{for } n \in \dots$ to combine in a way that results in an infinite loop.
  \end{itemize}
\end{itemize}

@@
